---
title: "Darwin Core mapping"
subtitle: "For: my_dataset_title"
author:
- author_1
- author_2
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
    number_sections: yes
    toc: yes
    toc_depth: 3
    toc_float: yes
---

<!-- For examples, see https://github.com/trias-project/checklist-recipe/wiki/Examples -->

# Setup 

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = TRUE)
```

Install required libraries (only if the libraries have not been installed before):

```{r}
installed <- rownames(installed.packages())
required <- c("tidyverse", "here", "janitor", "digest", "rgbif")
if (!all(required %in% installed)) {
  install.packages(required[!required %in% installed])
}
```

Load libraries:

```{r message = FALSE}
library(tidyverse)      # To do data science
library(here)           # To find files
library(janitor)        # To clean input data
library(digest)         # To generate hashes
library(rgbif)          # To use GBIF services
```

# Read source dat

Create a data frame `raw_data` from the source data. Please us the template `checklist.csv` provided in the `data/raw` folder.
We strongly advise to *not* read your data straight from an Excel spreadsheet due to encoding issues, but export the data to CSV format first.

```{r}
raw_data <- read_csv(here("data", "raw", "checklist.csv"), show_col_types = FALSE) 
```

Another option is to manage your data online, such as in a Google Sheet, and [publish the data to the web as a CSV](https://support.google.com/docs/answer/183965?hl=en&co=GENIE.Platform%3DDesktop). This will create an URL that can be used to read the data directly into R.

```{r}
# raw_data <- read_csv("MY_URL", col_types = cols(.default = "c"))
```

# Preprocessing

## Tidy data

Clean data somewhat:

```{r}
input_data <-
  raw_data |>
  remove_empty("rows") |>
  clean_names() 
```

## Check scientific names

Match the scientific names from your checklist against the GBIF Backbone Taxonomy, with the optional classification provided.

Column name | Description
-- | --
verbatim_scientific_name | The original scientific name as provided in the raw data file
scientificName | The matching scientific name from the GBIF backbone
matchType | The type of match (e.g. "EXACT", "FUZZY", "HIGHERRANK", "NOMATCH")
confidence | The confidence score of the match
rank | The taxonomic rank of scientificName
status | The taxonomic status of scientificName (e.g. "ACCEPTED", "SYNONYM", "DOUBTFUL")
acceptedUsageKey | The GBIF backbone taxonKey/usageKey/nubkey of the accepted name (if scientificName is a synonym)
acceptedScientificName | The accepted name (if scientificName is a synonym)



```{r}
input_data |>
  dplyr::mutate(
    scientificName = scientific_name,
    rank = taxon_rank
  ) |>
  name_backbone_checklist() |>
  dplyr::rename(rawScientificName = verbatim_scientific_name) |>
  dplyr::select(
    rawScientificName, scientificName, matchType, confidence, rank, status, acceptedUsageKey,
    acceptedScientificName
  )
```

Make sure to check the results of the name matching carefully, and correct any errors in the raw data file.

Once you upload your checklist to an Integrated Publishing Toolkit (IPT) and register it with GBIF, the system will automatically link the scientific names in your raw data file to the GBIF backbone taxonomy. 
The linking is based on the raw scientific name including authorship, the taxon rank and the provided taxonomy (eg kingdom). While GBIF keeps your raw data accessible on the published page, names that don't find a match in the GBIF Backbone won’t have active links to taxon web pages. To ensure your checklist is fully integrated and easy to navigate, your names should match as closely as possible with the GBIF Backbone.

Also consider to use the accepted taxon names instead of synonyms. GBIF automatically links synonyms with their accepted counterparts, but it’s best practice to use accepted taxon names directly in your checklist. It keeps things clean and prevents any unnecessary confusion for the users.

After you have made corrections, run all code again from the beginning.

## Generate taxon IDs

To link taxa with information in the extension(s), each taxon needs a unique and stable `taxonID`. Here we create one in the form of `dataset_shortname:taxon:hash`, where `hash` is unique code based on scientific name and kingdom. However, the `hash` will only remain the same as long as scientific name and kingdom remain the same. As soon as one character is changed, a new hash will be generated. Therefore, we strongly recommend to create the hash once for each taxon, and add it to the raw data file in the column `taxon_id_hash`. This way, the hash will remain stable even if scientific names change or are corrected in the future.

You can also provide your own id's, but make sure that if a taxon has multiple records, it has the same id in all records.

```{r}
vdigest <- Vectorize(digest) # Vectorize digest function to work with vectors
taxon_id_hashes <- 
    input_data |>
  mutate(
    # Only create a hash if taxon_id_hash is missing
    taxon_id_hash = if_else(
      is.na(taxon_id_hash),
      vdigest(paste(scientific_name, kingdom), algo = "md5"),
      taxon_id_hash
      )
  ) |>
  select(taxon_id_hash, scientific_name)

taxon_id_hashes
```

After you have filled in the `taxon_id_hash` column in the raw data file, run all code again from the beginning.

In the next code chunks, we remove the rows with missing `taxon_id_hash` values:

```{r}
(removed_rows <-
  input_data |>
  filter(is.na(taxon_id_hash)))

input_data <-
  input_data |>
  filter(!is.na(taxon_id_hash))
```

```{r}
cat(paste0(
  "Number of rows with missing `taxon_id_hash`: ", nrow(removed_rows),
  "\nNumbers of rows remaining in the checklist: ", nrow(input_data)
))
```

Next, we create the `taxon_id` column based on the `taxon_id_hash` column:

```{r}
input_data <-
  input_data |>
  mutate(
    taxon_id = paste(
      "my_dataset_shortname", # e.g. "alien-fishes-checklist"
      "taxon",
      taxon_id_hash,
      sep = ":"
    ))
```

# Darwin Core mapping

## Taxon core

Create a dataframe with unique taxa only (ignoring multiple distribution rows). Map the data to [Darwin Core Taxon](http://rs.gbif.org/core/dwc_taxon_2015-04-24.xml).

Start with record-level terms which contain metadata about the dataset (which is generally the same for all records). Make sure to replace the dummy values with your own values.
Then the taxon-level terms will be added in the next step.

```{r}
taxon <-
  input_data |>
  distinct(taxon_id, .keep_all = TRUE) |>
  mutate(
    # RECORD-LEVEL
    language = "my_language", # e.g. "en"
    license = "my_license", # e.g. "http://creativecommons.org/publicdomain/zero/1.0/"
    rightsHolder = "my_rights_holder", # e.g. "INBO"
    datasetID = "my_dataset_doi", # e.g. "https://doi.org/10.15468/xvuzfh"
    institutionCode = "my_institution_code", # e.g. "INBO"
    datasetName = "my_dataset_title", # e.g. "Checklist of non-native freshwater fishes in Flanders, Belgium"
    # TAXON-LEVEL
    taxonID = taxon_id,
    scientificName = scientific_name,
    kingdom = kingdom,
    phylum = phylum,
    class = class,
    order = order,
    family = family,
    genus = genus,
    taxonRank = taxon_rank,
    nomenclaturalCode = nomenclatural_code
  ) |>
  # Arrange by taxonID and only keep the Darwin Core columns
  arrange(taxonID) |> 
  select(
    "language", "license", "rightsHolder", "datasetID", "institutionCode",
    "datasetName", "taxonID", "scientificName", "kingdom", "phylum", "class",
    "order", "family", "genus",
    "taxonRank", "nomenclaturalCode"
  )
```

## Distribution extension

Create a dataframe with all data (including multiple distributions). Map the data to [Species Distribution](http://rs.gbif.org/extension/gbif/1.0/distribution.xml).

```{r}
distribution <-
  input_data |>
  # pathway
  pivot_longer(
    names_to = "key", 
    values_to = "pathway", 
    starts_with("introduction_pathway"),
    values_drop_na = FALSE) |>
  filter( # keep NA value for species with no pathway provided
    !is.na(pathway) |
      (is.na(pathway) & key == "introduction_pathway_1")
      ) |> 
  # other terms
  mutate(
    taxonID = taxon_id,
    locality = location,
    countryCode = country_code,
    occurrenceStatus = occurrence_status,
    establishmentMeans = establishment_means,
    degreeOfEstablishment = degree_of_establishment,
    eventDate = case_when(
      is.na(date_first_observation) & is.na(date_last_observation) ~ NA,
      is.na(date_first_observation) ~ paste0("/", date_last_observation),
      is.na(date_last_observation) ~ paste0(date_first_observation, "/"),
      !is.na(date_first_observation) & !is.na(date_last_observation) ~ 
        paste(date_first_observation, date_last_observation, sep = "/")
    ),
    source = source,
    occurrenceRemarks = occurrence_remarks
  ) |>
  # Arrange by taxonID and only keep the Darwin Core columns
  arrange(taxonID) |>
  select(
    "taxonID", "locality", "countryCode", "occurrenceStatus",
    "establishmentMeans", "degreeOfEstablishment", "pathway", "eventDate",
    "source", "occurrenceRemarks"
  ) 
```

## Species profile extension

In this extension we will express broad habitat characteristics of the species (e.g. `isTerrestrial`).

Create a dataframe with unique taxa only (ignoring multiple distribution rows).
Only keep records for which `terrestrial`, `marine` and `freshwater` is not empty.

Map the data to [Species Profile](http://rs.gbif.org/extension/gbif/1.0/speciesprofile.xml).

```{r}
species_profile <-
  input_data |>
  distinct(taxon_id, .keep_all = TRUE) |> 
  filter(
    !is.na(terrestrial) |
      !is.na(marine) |
      !is.na(freshwater)
  ) |> 
  mutate(
    .keep = "none",
    taxonID = taxon_id,
    isMarine = marine,
    isFreshwater = freshwater,
    isTerrestrial = terrestrial
  ) |> 
  arrange(taxonID)
```

## Description extension

In the description extension we want to include the native range of a species.

```{r}
description <-
  input_data |> 
  # unique taxa only (ignoring multiple distribution rows)
  distinct(taxon_id, .keep_all = TRUE) |> 
  # Separate values on `|` 
  mutate(native_range = strsplit(native_range, "\\|")) |> 
  unnest(native_range) |> 
  filter(!is.na(native_range)) |> 
  mutate(
    .keep = "none",
    taxonID = taxon_id,
    description = str_trim(native_range),
    type = "native range",
    language = "en"
    ) |> 
  arrange(taxonID)
```

# Save to CSV:

```{r}
write_csv(taxon, here("data", "processed", "taxon.csv"), na = "")
write_csv(distribution, here("data", "processed", "distribution.csv"), na = "")
write_csv(species_profile, here("data", "processed", "speciesprofile.csv"), na = "")
write_csv(description, here("data", "processed", "description.csv"), na = "")
```
